Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions

### 2022-05-25 SIMPaper/SIMPaper.tex
cleaned, licensed version of cbnn

---
title: "Case-Base Neural Networks: survival analysis with time-varying, higher-order interactions"
author:
- name: Jesse Islam
  num: a
- name: Maxime Turgeon
  num: b
- name: Robert Sladek
  num: c
- name: Sahir Bhatnagar
  num: d
address:
- num: a
  org: Department of Quantitative life sciences, McGill University 
- num: b
  org: Department of Statistics, University of Manitoba
- num: c
  org: Department of Human Genetics, McGill University
- num: d
  org: Department of Biostatistics, McGill University
corres: "\\email{jesse.islam@mail.mcgill.ca}"
authormark: Islam \emph{et al}.
articletype: Research article
received: 2022-04-29
revised: 2022-06-01
accepted: 2022-07-01
abstract: "In survival analysis, covariate interactions are difficult to model using regression-based methods without prior knowledge. Neural network-based survival models have been developed for data-driven interaction modeling, however, these approaches lack implementations for time-varying interactions or a flexible baseline hazard. To address this, we extend the case-base sampling framework and propose Case-Base Neural Networks (CBNN), which introduce time directly into the model, using commonly available neural network components for a simple implementation. First, we case-base sample the survival data, converting the survival time of each participant into discrete moments in time. This allows us to estimate the probability an event occurs at a given moment using a neural network to get the hazard. We compare CBNN to survival methods based on regression and neural networks in two simulations and two real data applications with a fixed neural network architecture. We report metrics for each model over follow-up time using both a proper (Index of Prediction Accuracy) and non-proper (inverse probability censoring weights-adjusted concordance index) scoring rule. A simple simulation shows CBNN outperforming the other neural network methods. The complex simulation highlights the ability of CBNN to model both a complex baseline hazard and time-varying interactions, outperforming all competitors. The first real data application shows CBNN outperforming all neural network competitors, while a second real data application shows competitive performance. CBNN has shown that it is more consistent in prediction performance across time in the simulations and real data applications compared to the competing neural network approaches."

keywords: survival analysis, machine learning, case-base, neural network
bibliography: bibfile.bib
output: 
  rticles::sim_article:
    includes:
      in_header: preamble.tex
    keep_tex: true
header-includes: 
  - \usepackage{placeins}
  - \newcommand{\sigmoid}{\mathrm{sigmoid}}
  - \renewcommand{\thefootnote}{\alph{footnote}}
---

In survival analysis, covariate interactions are difficult to model using regression-based methods without prior knowledge. Neural network-based survival models have been developed for data-driven interaction modeling, however, these approaches can not model both time-varying interactions and complex baseline hazards. To address this, we propose Case-Base Neural Networks (CBNN) which extends the case-base sampling framework which introduce time directly into the model. We case-base sample the survival data, converting the survival time of each participant into discrete moments in time. This allows us to use a feed-forward neural network that estimate the probability of an event occurring at a given moment, which can be used to estimate the hazard function. We then compare CBNN to survival methods based on regression and neural networks in two simulations and two real data applications with a fixed neural network architecture. We report metrics for each model over follow-up time using both proper (Index of Prediction Accuracy) and non-proper (inverse probability censoring weights-adjusted concordance index) scoring rules. A simple simulation with an exponential hazard model shows CBNN consistently outperforming the other neural network methods. The complex simulation highlights the ability of CBNN to model both a complex baseline hazard and time-varying interactions, outperforming all competitors. The first real data application shows CBNN outperforming all neural network competitors, while a second real data application shows competitive performance. In the simulations and real data applications, CBNN provides a more consistent predictive performance across time than the competing neural network approaches.

# Introduction

what is aft, dont like assuming distribution, smooth. Less used than cox. Proportional hazards assumptions hard to maintain in long study, all regression require prior knowledge.

The widespread use of the Cox proportional hazards model has led to a focus on survival models based on hazard ratios and relative risks rather than on survival curves and absolute risks [@hanley2009].
In clinical settings, Cox proportional hazards models are frequently prioritized over smooth-in-time accelerated failure time (AFT) models [@hanley2009], which can estimate absolute risks by modeling the hazard directly through user specification of a baseline hazard distribution. Using splines as the baseline hazard is an alternative to assuming a specific distribution [@royston2002flexible]. The risk associated with each covariate may not be constant over time, making it difficult to maintain the proportional hazards assumption. Previous studies related to breast cancer have discovered time-varying interactions with covariates of interest, like tumor size [@coradini2000time]. A limitation of regression-based models is the requirement of prior knowledge; the analyst must know potential time-varying interactions and explicitly evaluate them.

Smooth-in-time accelerated failure time (AFT) models can estimate absolute risks by modeling the hazard directly through user specification of a baseline hazard distribution [@kleinbaum2012survival]. The assumption of a specific distribution tends to be a point of contention. A method has been developped using splines, providing flexibility to the baseline hazard  [@royston2002flexible]. AFT models are underutilized compared to Cox proportional hazards, resulting in a focus on survival models based on hazard ratios and relative risks rather than on survival curves and absolute risks [@hanley2009]. However, The risk associated with each covariate may not be constant over time, making it difficult to maintain the proportional hazards assumption. Previous studies related to breast cancer have discovered time-varying interactions with covariates of interest, like tumor size [@coradini2000time]. A limitation of all regression-based models is the requirement of prior knowledge; the analyst must know potential time-varying interactions and explicitly evaluate them.

Neural networks provide a data-driven approach to approximating interaction terms. DeepSurv is a neural network-based proportional hazards model that implements the Cox partial log-likelihood as a custom loss function [@katzman2018DeepSurv]. This results in a stepwise absolute risk curve that cannot accommodate time-varying interactions. Compared to Cox regression, DeepSurv has shown an improvement in performance on the Study to Understand Prognoses Preferences and Risk Treatments (SUPPORT) dataset [@knaus1995SUPPORT]. Farraggi et al. suggested a modification to the loss function required for a Cox neural network that can handle non-proportional hazards [@faraggi1995neural]. Wang et al. took the concept of extreme learning machines and applied them in a survival context, removing the need for tuning the number of layers and nodes [@wang2019extreme].

DeepHit requires that each survival time of interest be specified in the model and directly estimates survival curves, rather than deriving a hazard function [@lee2018DeepHit]. It assumes an inverse Gaussian distribution as the baseline hazard and its concordance index (c-index) score outperforms DeepSurv [@katzman2018DeepSurv] on the Molecular Taxonomy of Breast Cancer International Consortium (METABRIC) dataset [@curtis2012genomic].

Deep Survival Machines (DSM) provides a flexible, parametric survival model using neural networks with a mixture of distributions as the baseline hazard [@dsmPaper]. On both the SUPPORT and METABRIC datasets, DSM outperforms DeepSurv and DeepHit [@dsmPaper]. However, like DeepHit and DeepSurv, DSM cannot model time-varying interactions.

These methods need custom loss and activation functions that are difficult to implement in practice. We note the complexity of these alternative neural network approaches; DeepSurv, DeepHit, and DSM all require custom loss functions [@katzman2018DeepSurv] [@lee2018DeepHit] [@dsmPaper], while DSM requires a two-phase learning process and user implementations for distributions beyond Log-Normal or Weibull [@dsmPaper]. Regression-based approaches require prior specification of all interaction terms, which makes it challenging to model covariate effects that change over time. In this article, we propose Case-Base Neural Networks (CBNN) as a method that models time-varying interactions and a flexible baseline hazard using commonly available neural network components. Our approach to modeling the full hazard uses case-base sampling [@hanley2009]. This sampling technique allows probabilistic models to predict survival outcomes. Following the case-base framework, we can use transformations of time as a feature (covariate) to specify different baseline hazards. For example, by including splines of time as covariates, we can approximate the Royston-Parmar flexible baseline hazard model [@royston2002flexible], however, this still requires user specification of time-varying interactions. CBNN can model both without user specification; only a sufficiently deep network is required.

In Section \ref{methods}, we describe how case-base sampling and neural networks are combined both conceptually and algebraically, along with our hyperparameter choices and software implementation. In Section \ref{sims}, we describe our metrics and compare the performance of CBNN, DeepSurv, DeepHit, DSM, Cox regression, and case-base using logistic regression on simulated data. Section \ref{casestudies} describes the real-data analysis, while in Section \ref{discussion}, we explore the implications of our results and contextualize them within neural network survival analysis in a single event setting.

# Case-base neural network methodology, metrics, and software {#methods}

In this section, we define case-base sampling, which converts the total survival time into discrete person-specific moments (person-moments). Then, we detail how neural networks can be used within this framework, explicitly incorporating time as a feature while adjusting for the sampling bias. Finally, we report on the software versions used. A package is available for use at [GITHUBLINK]. The entire code base to generate the analyses in this paper is available at [GITHUB LINK].

## Case-base sampling

Case-base sampling is an alternative framework for survival analysis [@hanley2009]. In case-base sampling, we sample from the continuous survival time of each person in our dataset to create a *base series* of *person-moments*. This *base series* complements the *case series*, which contains all person-moments at which the event of interest occurs.

For each person-moment sampled, let $X_i$ be the corresponding covariate profile $\left(x_{i1},x_{i2},...,x_{ip} \right)$, $T_i$ be the time of the person-moment, and $Y_i$ be the indicator variable for whether the event of interest occurred at time $T_i$. We estimate the hazard function $h(t \mid X_i)$ using the sampled person-moments. Recall that $h(t \mid X_i)$ is the instantaneous potential of experiencing the event at time $t$ for a given set of covariates $X_i$, assuming $T_i \geq t$.

Now, let $b$ be the (user-defined) size of the *base series*, and let $B$ be the sum of all follow-up times for the individuals in the study. If we sample the *base series* uniformly, then the hazard function of the sampling process is equal to $b/B$. Therefore, we have the following equality \footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}: \begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} 
The odds of a person-moment being a part of the *case series* is the ratio of the hazard $h(T_i \mid X_i)$ and the uniform rate $b/B$. Using \eqref{eqn:main}, we can see how the log-hazard function can be estimated from the log-odds arising from case-base sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

We may estimate the hazard function if we adjust for the bias introduced when sampling a fraction of the study base $B$ by adding the offset term $\log\left(\frac{b}{B} \right)$ (\eqref{eqn:offset}). Next, we propose using neural networks to model the odds.

## Neural networks to model the hazard function

@@

\begin{align}\label{eqn:nnProb}
P\left(Y=1|X,T\right)&=\sigmoid\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) \right),
\end{align}

where $T$ is a random variable representing the event time, $X$ is the random variable for a covariate profile, $f_{\theta}(X, T)$ represents any feed-forward neural network architecture, $\log\left(\frac{B}{b}\right)$ is bias term set by case-base sampling, $\theta$ is the set of parameters learned by the neural network and $\sigmoid(x)=\frac{1}{1+e^{-x}}$. By allowing a neural network to approximate a higher-order polynomial of time, the baseline hazard specification is now data-driven, where user-defined hyperparameters such as regularization, number of nodes, and layers control the flexibility of the model. We provide a detailed description of the choices we made in the next sub-section.

```{r NNarch,echo = FALSE, fig.align = 'center', out.width = "100%", fig.cap = "Steps involved in CBNN from case-base sampling to the model framework we use for training. The first step is case-base sampling, completed before training begins. Next, we pass this sampled data through a feed-forward neural network. We add the offset and pass that through a sigmoid activation function, whose output is a probability. Once the neural network model completes its training, we can convert the probability output to a hazard, using it for our survival outcomes of interest."}
library(knitr)
knitr::include_graphics(here::here("figures", "nnarch2.PNG"))
```

The following derivation shows how our probability estimate is converted to odds:
\begin{align*}
 \log\left( h(t \mid X_i) \right) &= \log\left(\frac{\sigmoid \left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{1-\sigmoid \left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left( \frac{\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}{1-\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left(\exp\left( f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) \right) \right) + \log\left(\frac{b}{B}\right) \\
 &= f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) + \log\left(\frac{b}{B}\right) \\
&= f_{\theta}(X, T). 
\end{align*}

We use binary cross-entropy as our loss function [@gulli2017]:
\begin{align*}
L(\theta)=-\frac{1}{N} \sum^{N}_{i=1} y_{i} \cdot \log(\hat{f}_{\theta}(x_{i}, t_{i}) ) + (1-y_{i} )\cdot \log(1-\hat{f}_{\theta}(x_{i}, t_{i}) ),
\end{align*}
where $\hat{f}_{\theta}(x_{i}, t_{i})$ is our estimate for a given covariate profile and time, $y_{i}$ is our target value specifying whether an event occurred, and $N$ represents the number of individuals in our training set.

Backpropagation is used to optimize the parameters in the model with an appropriate minimization algorithm (e.g. Adam, RMSPropagation, stochastic gradient descent)[@gulli2017]. For our analysis, we use Adam. Note that the size of the *case series* is fixed as the number of events, but we can make the *base series* as large as we want. A ratio of 100:1 *base series* to *case series* is sufficient [@hanley2009]. We pass our feed-forward neural network through a sigmoid activation function (Figure \ref{fig:NNarch}). Finally, we can convert this model output to a hazard. When using our model for predictions, we manually set the offset term to 0 in the new data, as we account for the bias during the fitting process.

Since we are directly modeling the hazard, we can readily estimate the risk function ($F$) at time $t$ for a covariate profile $X$, viz.\begin{align}\label{eqn:ci2}
F\left(t|X\right)& = 1 - \exp\left(-\int_{0}^{t}h(u|X) \,\textrm du\right).
\end{align} 
We use Riemann's Sum [@hughes2020calculus] to approximate the integral in \eqref{eqn:ci2}.

## Hyperparameter selection

Neural networks provide flexibility when defining the architecture and optimization parameters of our models. These hyperparameter choices can affect the estimated parameters and were chosen during a set of initial simulations to determine if CBNN can learn complex interactions in practice. We set a max number of epochs to be $2000$, batch size as $512$, learning rate as $10^{-3}$, decay as $10^{-7}$, patience as $10$ epochs, $\{50,50,25,25\}$ nodes in each hidden layer with $50\%$ dropout at each layer, a validation split of $20\%$, a testing split of $10%$, a minimum delta loss on the validation set of $10^{-7}$ over 10 epochs, and Adam [@gulli2017] as our optimizer. These choices may permit the model to approximate higher-order interactions while preventing over-fitting [@srivastava2014dropout]. We fix a train-validation-test split that allows us to update the weights with a subset of the data (training), assess performance at each epoch (validation) and gauge performance of the final model (test). We select the best weights after training based on validation loss [@gulli2017].

## Software implementation

R [@Rsoft] and python [@py] are used to evaluate methods from both languages. We fit the Cox model using the **survival** package [@survpkg], the CBLR model using the **casebase** package [@cbpkg], the DeepSurv model using the **survivalmodels** package [@survmods], the DSM model using **DeepSurvivalMachines** [@dsmPaper] and the DeepHit model using **pycox** [@lee2018DeepHit]. We made the components of CBNN using the **keras** [@keras] package and the **casebase** package for the sampling step. The **simsurv** package [@simsurv] is used to simulate our simulation studies, while **flexsurv** [@flexsurv] is used to fit a flexible baseline hazard using splines for our complex simulation. We use the implementation of c-ipcw from the python package **sksurv** [@sksurv]. The **riskRegression** package [@riskRegression] is used to get the Brier score and IPA metrics. We changed the **riskRegression** package to be used with any user supplied risk function $F$. To ensure that both R and Python-based models are running in unison on the same data through our simulations and bootstrap, we use the **reticulate** package [@reticulate].

# Simulation studies {#sims}

In this section, we use simulated data to evaluate the performance of CBNN and compare our approach with existing methods based on regression (Cox, CBLR) and neural network (DeepHit, DeepSurv, DSM) methods. We specify a linear combination of each covariate as the linear predictor in regression-based approaches (Cox, CBLR), which contrasts with neural network approaches that allow for non-linear interactions. We simulate data under a simple exponential model, and a complex baseline hazard with time-varying interactions, each with 10% random censoring. For both settings, we simulate three covariates:

$$
z_{1} \sim \textrm{Bernoulli}(0.5) \qquad \qquad
z_{2} \sim \begin{cases}
 N(0,0.5) & \textrm{if } z_{1}=0\\ 
 N(10,0.5) & \textrm{if } z_{1}=1
\end{cases} \qquad \qquad
z_{3} \sim \begin{cases}
 N(8,0.5) & \textrm{if } z_{1}=0\\ 
 N(-3,0.5) & \textrm{if } z_{1}=1.
\end{cases}
$$

>>>>

\documentclass[AMA,STIX1COL,]{WileyNJD-v2}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{longtable}
\newcommand{\srb}[1]{\textsf{\textcolor{red}{#1}}}
\usepackage{ulem}
\usepackage{placeins}
\newcommand{\sigmoid}{\mathrm{sigmoid}}
\renewcommand{\thefootnote}{\alph{footnote}}

\articletype{Research article}

\received{2022-04-29}

\revised{2022-06-01}

\accepted{2022-07-01}

\raggedbottom

\begin{document}

\title{Case-Base Neural Networks: survival analysis with time-varying,
higher-order interactions}

\author[a]{Jesse Islam}
\author[b]{Maxime Turgeon}
\author[c]{Robert Sladek}
\author[d]{Sahir Bhatnagar}

\authormark{Islam \emph{et al}.}

\address[a]{Department of Quantitative life sciences, McGill University}
\address[b]{Department of Statistics, University of Manitoba}
\address[c]{Department of Human Genetics, McGill University}
\address[d]{Department of Biostatistics, McGill University}

\corres{\email{jesse.islam@mail.mcgill.ca}}

\abstract{In survival analysis, covariate interactions are difficult to
model without prior knowledge using regression-based methods. Neural
network-based survival models have been developed for data-driven
interaction modeling, however, these approaches lack implementations for
time-varying interactions or a flexible baseline hazard. To address
this, we extend the case-base sampling framework and propose Case-Base
Neural Networks (CBNN), which introduce time directly into the model,
using commonly available neural network components for a simple
implementation. First, we case-base sample the survival data, converting
the survival time of each participant into discrete moments in time.
This allows us to estimate the probability an event occurs at a given
moment using a neural network to get the hazard. We compare CBNN to
survival methods based on regression and neural networks in two
simulations and two real data applications with a fixed neural network
architecture. We report metrics for each model over follow-up time using
both a proper (Index of Prediction Accuracy) and non-proper (inverse
probability censoring weights-adjusted concordance index) scoring rule.
A simple simulation shows CBNN outperforming the other neural network
methods. The complex simulation highlights the ability of CBNN to model
both a complex baseline hazard and time-varying interactions,
outperforming all competitors. The first real data application shows
CBNN outperforming all neural network competitors, while a second real
data application shows competitive performance. CBNN has shown that it
is more consistent in prediction performance across time in the
simulations and real data applications compared to the competing neural
network approaches.}

\keywords{survival analysis, machine learning, case-base, neural
network}

\maketitle

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

The widespread use of the Cox proportional hazards model has led to a
focus on survival models based on hazard ratios and relative risks
rather than on survival curves and absolute risks \citep{hanley2009}. In
clinical settings, Cox models are frequently prioritized over
smooth-in-time accelerated failure time (AFT) models \citep{hanley2009},
which can estimate absolute risks by modeling the hazard directly
through user specification of a baseline hazard distribution. The user
may forgo this choice by implementing a flexible set of parameters using
splines, as proposed by Royston and Palmar \citep{royston2002flexible}.
These regression models also require user-specified interactions, which
may limit risk prediction accuracy.

Over longer follow-up, it is challenging to maintain the proportional
hazards assumption as the risk associated with each covariate may not be
constant with time. Previous studies related to breast cancer have
discovered time-varying interactions with covariates of interest, like
tumor size \citep{coradini2000time} \citep{hilsenbeck1998time}. Another
limitation of regression-based models is the requirement of prior
knowledge; the analyst must know potential time-varying interactions and
explicitly evaluate them.

Data-driven neural network approaches bypass the need for prior
knowledge of interaction terms. DeepSurv is a semi-parametric
proportional hazards neural network model that implements the Cox
partial log-likelihood as a custom loss function
\citep{katzman2018DeepSurv}. This results in a stepwise absolute risk
curve that cannot accommodate time-varying interactions. Compared to Cox
regression, DeepSurv has shown a small improvement in performance using
a single layer neural network on the Study to Understand Prognoses
Preferences and Risk Treatments (SUPPORT) dataset
\citep{knaus1995SUPPORT}. Farraggi et al.~suggested a modification to
the loss function required for a Cox neural network that can model
non-proportional hazards \citep{faraggi1995neural}. Wang et al.~took the
concept of extreme learning machines and applied them in a survival
context, removing the need for tuning the number of layers and nodes
\citep{wang2019extreme}. A Cox-based neural network combined with Local
Interpretable Model-Agnostic Explanations (LIME)
\citep{ribeiro2016model} interprets the association of each Single
Nucleotide Polymorphism (SNP) \citep{sun2020genome} to a phenotype.

DeepHit requires each survival time of interest to be specified in the
model and directly estimates survival curves, rather than deriving a
hazard function \citep{lee2018DeepHit}. It assumes an inverse Gaussian
distribution as the baseline hazard and its concordance index (c-index)
score outperforms DeepSurv \citep{katzman2018DeepSurv} on the Molecular
Taxonomy of Breast Cancer International Consortium (METABRIC) dataset
\citep{curtis2012genomic}.

Ripley et al.~presented seven different distributions and modeling
paradigms based on neural networks \citep{ripley2004non}. They provided
the methodology behind parametric log-logistic and log-normal neural
network models. Like DeepHit, these models are restricted to one of the
seven baseline hazard distributions, reducing the flexibility of the
average survival prediction.

Deep Survival Machines (DSM) provides a flexible, parametric survival
model using neural networks with a mixture of distributions as the
baseline hazard \citep{dsmPaper}. DSM allows the user to specify a
mixture of distributions for a flexible baseline hazard. On both the
SUPPORT and METABRIC datasets, DSM outperforms DeepSurv and DeepHit
\citep{dsmPaper}. However, like DeepHit and DeepSurv, DSM cannot model
time-varying interactions.

These methods need custom loss and activation functions that are
difficult to implement in practice. We note the complexity of these
alternative neural network approaches; DeepSurv, DeepHit, and DSM all
require custom loss functions \citep{katzman2018DeepSurv}
\citep{lee2018DeepHit} \citep{dsmPaper}, while DSM requires user
implementations for distributions beyond Log-Normal or Weibull and a
two-phase learning process \citep{dsmPaper}. Regression-based approaches
require prior specification of all interaction terms, which makes it
challenging to model covariate effects that change over time. In this
article, we propose Case-Base Neural Networks (CBNN) as a method that
models time-varying interactions and a flexible baseline hazard using
commonly available neural network components. Our approach to modeling
the full hazard uses case-base sampling \citep{hanley2009}, which allows
probabilistic models to predict survival outcomes. Following the
case-base framework, we can use transformations of time as a feature
(covariate) to specify different baseline hazards. For example, by
including splines of time as covariates, we can approximate the
Royston-Parmar flexible baseline hazard model
\citep{royston2002flexible}. An alternative approach to Royston-Parmar
is case-base with logistic regression (CBLR), which provides some
flexibility in modeling, but still requires user specification of
time-varying interactions and baseline hazard, while CBNN can model both
without user specification; only a sufficiently deep network is
required.

In Section \ref{methods}, we describe how case-base sampling and neural
networks are combined both conceptually and algebraically, along with
our hyperparameter choice and software implementation. Then, in Section
\ref{sims}, we describe our metrics and compare the performance of CBNN,
DeepSurv, DeepHit, DSM, Cox regression, and case-base using logistic
regression on simulated data. Next, Section \ref{casestudies} describes
the real-data analysis, while in Section \ref{discussion}, we explore
the implications of our results and contextualize them within neural
network survival analysis in a single event setting.

\hypertarget{methods}{%
\section{Case-base neural network methodology, metrics, and
software}\label{methods}}

In this section, we define case-base sampling, which converts the total
survival time into discrete person-specific moments (person-moments).
Then, we detail how neural networks can be used within this framework,
explicitly incorporating time as a feature while adjusting for the
sampling bias. Finally, we report on the software versions used. A
package is available for use at {[}GITHUBLINK{]}. The entire code base
to generate the analyses in this paper is available at {[}GITHUB
LINK{]}.

\hypertarget{case-base-sampling}{%
\subsection{Case-base sampling}\label{case-base-sampling}}

Case-base sampling is an alternative framework for survival analysis
\citep{hanley2009}. In case-base sampling, we sample from the continuous
survival time of each person in our dataset to create a \emph{base
series} of \emph{person-moments}. This \emph{base series} complements
the \emph{case series}, which contains all person-moments at which the
event of interest occurs.

For each person-moment sampled, let \(X_i\) be the corresponding
covariate profile \(\left(x_{i1},x_{i2},...,x_{ip} \right)\), \(T_i\) be
the time of the person-moment, and \(Y_i\) be the indicator variable for
whether the event of interest occurred at time \(T_i\). We estimate the
hazard function \(h(t \mid X_i)\) using the sampled person-moments.
Recall that \(h(t \mid X_i)\) is the instantaneous potential of
experiencing the event at time \(t\) for a given set of covariates
\(X_i\), assuming \(T_i \geq t\).

Now, let \(b\) be the (user-defined) size of the \emph{base series}, and
let \(B\) be the sum of all follow-up times for the individuals in the
study. If we sample the \emph{base series} uniformly, then the hazard
function of the sampling process is equal to \(b/B\). Therefore, we have
the following equality
\footnote{We are abusing notation here, conflating hazards with probabilities. For a rigorous treatment, see Saarela \& Hanley (2015) section 3 \cite{saarela2015} .}:
\begin{align}\label{eqn:main}
\frac{P\left(Y_i=1 \mid X_i, T_i\right)}{P\left(Y_i = 0 \mid X_i, T_i\right)} = \frac{h\left(T_i \mid X_i\right)}{b/B}.
\end{align} The odds of a person-moment being a part of the \emph{case
series} is the ratio of the hazard \(h(T_i \mid X_i)\) and the uniform
rate \(b/B\). Using \eqref{eqn:main}, we can see how the log-hazard
function can be estimated from the log-odds arising from case-base
sampling: \begin{align}\label{eqn:offset}
\log \left( h\left(t \mid X_i\right)\right) = \log \left(\frac{P\left(Y_i = 1 \mid X_i, t\right)}{P\left(Y_i = 0 \mid X_i, t\right)}\right) + \log\left(\frac{b}{B}\right).
\end{align}

We may estimate the hazard function if we adjust for the bias introduced
when sampling a fraction of the study base \(B\) by adding the offset
term \(\log\left(\frac{b}{B} \right)\) (\eqref{eqn:offset}). Next, we
propose using neural networks to model the odds.

\hypertarget{neural-networks-to-model-the-hazard-function}{%
\subsection{Neural networks to model the hazard
function}\label{neural-networks-to-model-the-hazard-function}}

@@

\begin{align}\label{eqn:nnProb}
P\left(Y=1|X,T\right)&=\mathrm{sigmoid}\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) \right),
\end{align}

where \(T\) is a random variable representing the event time, \(X\) is
the random variable for a covariate profile, \(f_{\theta}(X, T)\)
represents any feed-forward neural network architecture,
\(\log\left(\frac{B}{b}\right)\) is bias term set by case-base sampling,
\(\theta\) is the set of parameters learned by the neural network and
\(\mathrm{sigmoid}(x)=\frac{1}{1+e^{-x}}\). By allowing a neural network
to approximate a higher-order polynomial of time, the baseline hazard
specification is now data-driven, where user-defined hyperparameters
such as regularization, number of nodes, and layers control the
flexibility of the model. We provide a detailed description of the
choices we made in the next sub-section.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{/home/bhatnagar-lab/jislam/pmnnProjects/paper/figures/nnarch2}

}

\caption{Steps involved in CBNN from case-base sampling to the model framework we use for training. The first step is case-base sampling, completed before training begins. Next, we pass this sampled data through a feed-forward neural network. We add the offset and pass that through a sigmoid activation function, whose output is a probability. Once the neural network model completes its training, we can convert the probability output to a hazard, using it for our survival outcomes of interest.}\label{fig:NNarch}
\end{figure}

The following derivation shows how our probability estimate is converted
to odds: \begin{align*}
 \log\left( h(t \mid X_i) \right) &= \log\left(\frac{\mathrm{sigmoid}\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{1-\mathrm{sigmoid}\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left( \frac{\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}{1-\frac{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)}{\exp\left(f_{\theta}(X, T) + \log\left(\frac{B}{b}\right)\right)+1}}\right) + \log\left(\frac{b}{B}\right) \\
 &= \log\left(\exp\left( f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) \right) \right) + \log\left(\frac{b}{B}\right) \\
 &= f_{\theta}(X, T) + \log\left(\frac{B}{b}\right) + \log\left(\frac{b}{B}\right) \\
&= f_{\theta}(X, T). 
\end{align*}

We use binary cross-entropy as our loss function \citep{gulli2017}:
\begin{align*}
L(\theta)=-\frac{1}{N} \sum^{N}_{i=1} y_{i} \cdot \log(\hat{f}_{\theta}(x_{i}, t_{i}) ) + (1-y_{i} )\cdot \log(1-\hat{f}_{\theta}(x_{i}, t_{i}) ),
\end{align*} where \(\hat{f}_{\theta}(x_{i}, t_{i})\) is our estimate
for a given covariate profile and time, \(y_{i}\) is our target value
specifying whether an event occurred, and \(N\) represents the number of
individuals in our training set.

Backpropagation is used to optimize the parameters in the model with an
appropriate minimization algorithm (e.g.~Adam, RMSPropagation,
stochastic gradient descent)\citep{gulli2017}. For our analysis, we use
Adam. Note that the size of the \emph{case series} is fixed as the
number of events, but we can make the \emph{base series} as large as we
want. A ratio of 100:1 \emph{base series} to \emph{case series} is
sufficient \citep{hanley2009}. We pass our feed-forward neural network
through a sigmoid activation function (Figure \ref{fig:NNarch}).
Finally, we can convert this model output to a hazard. When using our
model for predictions, we manually set the offset term to 0 in the new
data, as we account for the bias during the fitting process.

Since we are directly modeling the hazard, we can readily estimate the
risk function (\(F\)) at time \(t\) for a covariate profile \(X\),
viz.\begin{align}\label{eqn:ci2}
F\left(t|X\right)& = 1 - \exp\left(-\int_{0}^{t}h(u|X) \,\textrm du\right).
\end{align} We use Riemann's Sum \citep{hughes2020calculus} to
approximate the integral in \eqref{eqn:ci2}.

\hypertarget{hyperparameter-selection}{%
\subsection{Hyperparameter selection}\label{hyperparameter-selection}}

Neural networks provide flexibility when defining the architecture and
optimization parameters of our models. These hyperparameter choices can
affect the estimated parameters and were chosen during a set of initial
simulations to determine if CBNN can learn complex interactions in
practice. We set a max number of epochs to be \(2000\), batch size as
\(512\), learning rate as \(10^{-3}\), decay as \(10^{-7}\), patience as
\(10\) epochs, \(\{50,50,25,25\}\) nodes in each hidden layer with
\(50\%\) dropout at each layer, a validation split of \(20\%\), a
testing split of \(10%
\), a minimum delta loss on the validation set of \(10^{-7}\) over 10
epochs, and Adam \citep{gulli2017} as our optimizer. These choices may
permit the model to approximate higher-order interactions while
preventing over-fitting \citep{srivastava2014dropout}. We fix a
train-validation-test split that allows us to update the weights with a
subset of the data (training), assess performance at each epoch
(validation) and gauge performance of the final model (test). We select
the best weights after training based on validation loss
\citep{gulli2017}.

\hypertarget{software-implementation}{%
\subsection{Software implementation}\label{software-implementation}}

R \citep{Rsoft} and python \citep{py} are used to evaluate methods from
both languages. We fit the Cox model using the \textbf{survival} package
\citep{survpkg}, the CBLR model using the \textbf{casebase} package
\citep{cbpkg}, the DeepSurv model using the \textbf{survivalmodels}
package \citep{survmods}, the DSM model using
\textbf{DeepSurvivalMachines} \citep{dsmPaper} and the DeepHit model
using \textbf{pycox} \citep{lee2018DeepHit}. We made the components of
CBNN using the \textbf{keras} \citep{keras} package and the
\textbf{casebase} package for the sampling step. The \textbf{simsurv}
package \citep{simsurv} is used to simulate our simulation studies,
while \textbf{flexsurv} \citep{flexsurv} is used to fit a flexible
baseline hazard using splines for our complex simulation. We use the
implementation of c-ipcw from the python package \textbf{sksurv}
\citep{sksurv}. The \textbf{riskRegression} package
\citep{riskRegression} is used to get the Brier score and IPA metrics.
We changed the \textbf{riskRegression} package to be used with any user
supplied risk function \(F\). To ensure that both R and Python-based
models are running in unison on the same data through our simulations
and bootstrap, we use the \textbf{reticulate} package
\citep{reticulate}.

\hypertarget{sims}{%
\section{Simulation studies}\label{sims}}

In this section, we use simulated data to evaluate the performance of
CBNN and compare our approach with existing methods based on regression
(Cox, CBLR) and neural network (DeepHit, DeepSurv, DSM) methods. We
specify a linear combination of each covariate as the linear predictor
in regression-based approaches (Cox, CBLR), which contrasts with neural
network approaches that allow for non-linear interactions. We simulate
data under a simple exponential model, and a complex baseline hazard
with time-varying interactions, each with 10\% random censoring. For
both settings, we simulate three covariates:
